Validation of Visual Statistical Inference, Applied to Linear Models

### 2013-04-04 paper_jasa.tex
tiny change
conclusion is reduced

Visual inference falls into the same framework as randomization tests, where the statistics from the data are compared with those from null data. But unlike these approaches visual inference is constrained to make the comparison with just a few draws $(m-1)$ from the null distribution. How this small set of null plots influences the subjects choice is important for understanding the reliability of visual inference. If the actual data plot is very different from the null plots, then the null plots should not have much influence on the choice. Measuring the difference, generally, between  plots is almost impossible. However,  in this controlled setting we can use $p$-values of the test statistic calculated on the data used in each plot as a proxy for similarity of structure between the plots. If there is a null plot with a small $p$-value, or one close to that of the actual data plot, we would expect that subjects have a harder time detecting the actual data plot. The detailed discussion is attached to the supplementary material.

@@

@@

% when evaluating the lineup: this question can be lined up with the alternative hypothesis of a classical testing scenario. If we are interested in the reason why participants picked a plot as the most different, we need to collect and evaluate this information in the study. This has to be done very carefully, as any suggestions for reasons also lead the observer  what patterns to look out for.}
% By delivering an appropriate question may reduce the type III error in an appropriate scenario.}

@@

>>>>

Visual inference falls into the same framework as randomization tests, where the statistics from the data are compared with those from null data. But unlike these approaches visual inference is constrained to make the comparison with just a few draws $(m-1)$ from the null distribution. How this small set of null plots influences the subjects choice is important for understanding the reliability of visual inference. If the actual data plot is very different from the null plots, then the null plots should not have much influence on the choice. Measuring the difference, generally, between  plots is almost impossible. However,  in this controlled setting we can use $p$-values of the test statistic calculated on the data used in each plot as a proxy for similarity of structure between the plots. If there is a null plot with a small $p$-value, or one close to that of the actual data plot, we found that subjects have a harder time detecting the actual data plot. The detailed discussion is attached to the supplementary material.

@@

This paper has demonstrated that statistical graphics can be used in statistical inference and validates the lineup protocol proposed by \citet{buja:2009} as a formal framework to accomplish that. For this, specific terminologies were defined. The methods of obtaining $p$-value and estimating power of visual tests were introduced. If it is assumed that observers tend to pick the plot with the strongest signal, which was consistent with the experimental data, the power of visual test can be obtained theoretically.  Interestingly, the theoretical power of visual test is better than that of conventional test in the conditions tested in this paper. The simulation study showed that even in the situation where the conventional test has its assumptions met, visual inference is comparable. The results also suggest that when the underlying assumptions of a conventional test do not hold, visual inference will perform well. Even though a single observer can evaluate a lineup, as suggested in the initial lineup protocol, this work recommends having multiple observers to ensure enough power.

When there is no conventional test available, visual testing procedure can be used due to its few assumptions and non-parametric nature.  Lineups are successfully used to evaluate the power of competing graphical designs in \cite{heike:2012}. This method is also applied in \citet{niladri:2012}. It is hoped that this proves to be extremely valuable in data mining applications, and exploratory analyses, where there are no existing gauges of statistical significance. The lineup protocol operates similarly to general tests, such as two-sided, analysis of variance $F$-tests, and Wilks $\Lambda$ for multivariate analysis of variance. If the null hypothesis is rejected, generally we can say that ``there is something there'' but not specifically what it is in the data that triggers the rejection. Follow-up questions on the reasons provide qualitative insight. Multiple comparisons are often done to refine and understand the results of the general conventional testing, and perhaps some similar approaches might be developed for visual inference.

%One of the critical parts in visual testing is appropriate selection of a statistic for the hypothesis of interest -- that means an appropriate choice of plot for the analytical purpose.  A good test statistic is expected to reveal structure if the null hypothesis is not true. Small differences in plot choices such as color, orientation, and mapping of data values to graphical elements of the plot can make a difference. This is related to research in a grammar of graphics \citep{wilkinson:1999,hadley:2009}, which helps to ensure good practices.

@@

 %This is the part of our ongoing and future research.

%The question that is posed to observers can affect the ability to evaluate lineups. In these experiments subjects were asked very specific questions, because the purpose was to compare with very specific conventional testing. In practice, the question posed will be more general to allow observers to choose plots with any remarkable features, which enables the data mining discovery process.

% when evaluating the lineup: this question can be lined up with the alternative hypothesis of a classical testing scenario. If we are interested in the reason why participants picked a plot as the most different, we need to collect and evaluate this information in the study. This has to be done very carefully, as any suggestions for reasons also lead the observer  what patterns to look out for.}

% By delivering an appropriate question may reduce the type III error in an appropriate scenario.}

%Experimental subjects in the simulation study came from various level of education, both genders and many different ages. Many subjects had little background in statistics, and were not necessarily aware of statistical plots. Yet, the results were good. We expect that visual inference could be a useful tool for high school students to learn about randomness as a primer to statistical thinking. A major advantage of visual statistical inference is that it is intuitive, as illustrated by the performance of subjects with limited education, and it  may make statistics fun to learn.

@@

\section{Old Conclusions}

% when evaluating the lineup: this question can be lined up with the alternative hypothesis of a classical testing scenario. If we are interested in the reason why participants picked a plot as the most different, we need to collect and evaluate this information in the study. This has to be done very carefully, as any suggestions for reasons also lead the observer  what patterns to look out for.}

% By delivering an appropriate question may reduce the type III error in an appropriate scenario.}
